\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-17 k-planar.tex
Cleaned up and filled in an important missing detail

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, a $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

@@

\note{PM}{I have significantly simplified this section, which should make the referees happy. I will read it very carefully again tomorrow}

We begin with a standard technique that allows us to work with a normalised tree-decomposition whose tree has same vertex set as the graph it decomposes:

@@

  Next, remove the node $x_0$ from $T_0$, and consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.
  % To see that $(B_x:x\in V(T))$ satisifies \tref{ancestor-edge}, observe that,
  % if $xy\in E(H)$, then at least one of $x$ or $y$ is contained in $B_z$ for every node $z$ on the path from $x$ to $y$ in $T$.  If neither $x$ nor $y$ is an ancestor of the other, then some node $z$ on this path has $T$-depth less than that of $x$ and $y$.  If $x\in B_z$ this contradicts the fact that $x$ is the root of $T[x]$.  If $y\in B_z$ this contradicts the fact that $y$ is the root of $T[y]$.
\end{proof}

@@

%\subsection{Generalized Tripod Partitions}
%
%
% \begin{figure}[htbp]
%   \begin{center}
%     \includegraphics{figs/tripoddo}
%   \end{center}
%   \caption{The sets $Y_x$, $F_x$, and $V_x$ associated with $x\in V(T)$
%   and the ancestors $a_1,\ldots,a_{t'}$ of $X$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.}
%   \label{fig:generalized-tripod}
% \end{figure}

% The following lemma shows how to interpret an $H$-partition of $G$ and a tree-decomposition of $H$ as a `hierarchical' decomposition of $G$; refer to \cref{fig:generalized-tripod}.

% For a vertex $v$ of $G$, $N_G(v):=\{w\in V(G):vw\in E(G)\}$ and $N_G[v]:=N_G(v)\cup\{v\}$.  For a vertex set $S\subseteq V$, $N_G(S):=\bigcup_{v\in S} N_G(S)\setminus S$ and $N_G[S]:=N_G(S)\cup S$.

% \note{PM}{
% Guide to old notation:
% \begin{compactitem}
%   \item $V_x := \bigcup_{y\in V(T_x)} Y_y = G_x$
%   \item $F_x:=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}= N_G(G_x)$
%   \item $N_x:=V_x\cup F_x=N_G[G_x]$
% \end{compactitem}
% }
%
% \begin{lem}\label{generalized-tripod}
%   Let $G$ be a graph; let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$ where $H$ has treewidth at most $t$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.
%   For each $x\in V(T)$, let $H_x:=V(T_x)$ and let $G_x:=\bigcup_{y\in H_x}Y_y$.
%
%   \begin{compactenum}[(Y1)]
%     % \item\ylabel{thickness} $\mathcal{Y}=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width at most $\ell$ with respect to $\mathcal{L}$.
%     % \item\ylabel{separator} For each $x\in V(T)$, $N_G(\phi(T_x))$ separates $\phi(T_x)$ from $N
%     % there is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$.
%
%     \item\ylabel{ancestor-edge} For each $x\in V(T)$, there is a set $\{a_1,\ldots,a_{t'}\}$ of $t'\le t$ strict $T$-ancestors of $x$ such that $N_G(G_x) \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
%   \end{compactenum}
% \end{lem}
%
% \begin{proof}
%   Let $\{a_1,\ldots,a_{t'}\}:= B_x\setminus\{x\}$.  Then, by the definition of a normalized tree decomposition $N_H(H_x))\subseteq\{a_1,\ldots,a_{t'}\}$.  Therefore $N_G(G_x)\subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
% \end{proof}

% \referee{1}{1. Lemma 3: This lemma is really about ``normalised'' tree-decompositions, and says that for each $x \in V(T)$, the set $V(T_x)$ has at most $t$ neighbours in $H$. Call this property (T3) and remove (Y1) through (Y5), which should now be obvious.}
%
% \note{PM}{I've added to the \texttt{response.tex} file explaining some of this.  Each of (Y1), (Y3), (Y4), and (Y5) have a one-sentence justification. The justification for (Y2) is just definition unpacking and then using the observation that the referee recommends.}
%
% \note{PM}{I plan to do an experiment with the referee's suggested notation, in a separate git branch to see how it turns out.}
%
% Before proving \cref{generalized-tripod} we point out more properties that are immediately implied by it:
%
% \begin{compactenum}[(Y1)]\setcounter{enumi}{2}
%   \item\ylabel{y-subsets} $Y_x\subseteq V_x$ for every $x\in V(T)$. \note{PM}{$Y_x$, for $x\in V(T)$ is not defined!}
%   \item\ylabel{containment-i} $V_x\subseteq V_a$ for every $T$-ancestor $a$ of $x$.
%   \item\ylabel{containment-ii}$N_x\subseteq N_a$ for every $T$-ancestor $a$ of $x$.
% \end{compactenum}

% Property~\yref{y-subsets} follows from the fact that $V_x$ is the union of several sets, one of which is $Y_x$.  Property~\yref{containment-i} follows from the definition of $V_x$ and the fact that $V(T_x)\subseteq V(T_a)$. To show Property~\yref{containment-ii} first note that, by \yref{containment-i} it suffices to consider vertices $w\in F_x=N_x\setminus V_x$. By definition, every vertex $w\in F_x$ is adjacent, in $G$, to a vertex $v\in V_x$.  By \yref{containment-i}, $v\in V_a$, so $w$ is either in $V_a$ or $w$ satisfies the condition $vw\in E(G)$, $v\in V_a$, and $w\not\in V_a$, so $w\in F_a$.  In either case $w\in N_a=V_a\cup F_a$.  Note that none of \yref{y-subsets}--\yref{containment-ii} depends on \yref{ancestor-edge} (which is important, since \yref{containment-i} is used to establish \yref{ancestor-edge} in the following proof).
%
%
% \begin{proof}[Proof of \cref{generalized-tripod}]
%   % Property~\yref{thickness} follows immediately from the fact that $V(T)=V(H)$  and the fact that $\mathcal{Y}$ has layered width at most $\ell$ with respect to $\mathcal{L}$.
%   Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.
%
%   To establish Property~\yref{ancestor-edge}, consider some vertex $w\in F_x$.  Since $w\in F_x$, there exists an edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  Since $v\in V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\not\in V(T_x)$.  Since $vw\in E(G)$, we have $x'a\in E(H)$.  By \tref{ancestor-edge}, one of $a$ or $x'$ is a $T$-ancestor of the other. Since $w\in Y_a\subseteq V_a$ and $w\not\in V_x\supseteq V_{x'}$, \yref{containment-i} rules out the possibility that $x'$ is a $T$-ancestor of $a$. Therefore, $a$ is a $T$-ancestor of $x$ which is a $T$-ancestor of $x'$.  Let $z_0,\ldots,z_r$ be the path in $T$ from $z_0:=x'$ to $z_r:=a$.  Since $x'a\in E(H)$, at least one of $a$ or $x'$ is in $B_{z_i}$, for each $i\in\{0,\ldots,r\}$ and there is at least one $i$ such that $B_{z_i}$ contains both $x'$ and $a$.  However, by \tref{subtree-root} $x'$ is not contained in $B_{z_i}$ for any $i\in\{1,\ldots,r\}$.  Therefore $a\in B_{z_i}$ for each $i\in\{0,\ldots,r\}$.  In particular, $a$ is contained in $B_x$.
%   Property~\yref{ancestor-edge} now follows from the fact that $|B_x|\le t+1$ and $B_x$ contains $x$.
% \end{proof}

% \referee{2}{Page 7.  In the very last sentence, I think $B_{x_i}$ should be $B_{z_i}$
% (twice).  Moreover, the way the proof is written, it seems to suggest
% that we only get that $a \in B_{z_i}$ for each $i \in \{1, \dots, r\}$.
% The conclusion that $a \in B_{z_i}$ for each $i \in \{0, \dots, r\}$ is
% true though.  In a normalized tree-decomposition, if $xy$ is an edge and
% $x$ is a $T$-ancestor of $y$, then $x$ must be in $B_y$ (since the subtrees $T_x$
% and $T_y$ intersect).}

%\note{PM}{I addressed the preceding referee's.}

@@

% \referee{1}{2. Proof of Theorem 9: There is too much notation throughout. As
% 	one example, it is hard to remember which of $T_x$, $Y_x$, $V_x$, $F_x$, $N_x$,
% 	$S_x$, $X_v$, $B_x$, $C_x$ is which. If $\phi(S)$ is written for the subset of $V(G)$
% 	corresponding to a set $S \subseteq V(H)$ and appropriate shorthand and
% 	standard notation for neighbourhoods are used, then $Y_x$, $V_x$, $F_x$, and
% 	$N_x$ can be replaced by $\phi(x)$, $\phi(T_x)$, $N(\phi(T_x))$, and $V(G) \setminus N[\phi(T_x)]$.
% 	The set $X_v$ is only used in the statement of Claim~1, so it does not
% 	need a name; say what these vertices are explicitly. It may be more
% 	indicative to write $\phi'(x)$ for $S_x$ and $B'_x$ for $C_x$.}

\begin{proof}
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.
% For each $x\in V(T)$, let $H_x:=V(T_x)$ and let $G_x:=\bigcup_{y\in H_x}Y_y$.

% Apply \cref{generalized-tripod} to $G$ and let $\mathcal{L}$, $\mathcal{Y}$, $\mathcal{T}$, $T$, $H_x$, and $G_x$ be defined as in \cref{generalized-tripod}, where the partition $\mathcal{Y}$ has width $\ell$ with respect to the layering $\mathcal{L}$.

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$, or $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

% \begin{clm}
%   For each $v\in V(G)$, $v$ participates in $a(v)$.
% \end{clm}
%
% \begin{proof}
%   Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.
%
%   We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.
%
%   Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  Therefore $X_H$ contains every every node in which $v$ participates.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected. The fact that $\mathcal{T}$ is a normalized tree decomposition of $H$ implies that some node $a\in X_H$ is a $T$-ancestor of every node in $X_H$.  Thus, $v$ participates in the node $a=a(v)$ that is the lowest common $T$-ancestor of every node in which $v$ participates, as required.
% \end{proof}

% \begin{proof}
%   Let $a(v)$.  If $v\in Y_{a(v)}$ then $v$ participates in $a(v)$, by definition, and we are done.  Otherwise, there are two nodes $x,y\in V(T)$ in which $x$ participates and whose lowest common $T$-ancestor is $a(v)$.
%
%   There are two cases to consider:
%   \begin{compactenum}
%     \item If $v\in Y_x$ then $v\not\in Y_y$.  Since $v$ participates in $Y_y$, there exist a shortcut $P:=v_0,\ldots,v_r$ in $\SS$ with $v=v_\alpha$ and $v_\beta\in Y_y$ for some $\alpha\in\{0,\ldots,r\}$ and $\beta\in\{1,\ldots,r-1\}$.  Without loss of generality, $\alpha < \beta$.
%
%     For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  Since $G[\{v_\alpha,\ldots,v_\beta\}]$ is connected, $H[\{x_\alpha,\ldots,x_\beta\}]$ connected.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, $x_j$ is a $T$-ancestor of every node in $x_\alpha,\ldots,x_\beta$ for some $j\in\{\alpha,\ldots,\beta\}$.      In particular $x_j$ is a $T$-ancestor of $x=x_\alpha$ and $y=x_\beta$, so $x_j$ is a $T$-ancestor of $a(v)$.  Since $v_j\in V(P)$ and $v_j\in Y_{x_j}$ this implies that $v$ participates in $x_j$.  Therefore $v$ participates in a $T$-ancestor of $a(v)$, which can only mean that $v$ participates in $a(v)$.
%
%     \item If $v\not\in Y_x$ and $v\not\in Y_y$ then there are shortcuts $P_1,P_2\in\SS$ with $v\in V(P_1)$, $v\in V(P_2)$ where $P_1$ has an internal vertex $w_1\in Y_x$ and $P_2$ has an internal vertex $w_2\in Y_y$.
%
%
%     .  Since $v$ participates in $Y_y$, there exist a shortcut $v_0,\ldots,v_r$ in $\SS$ with $v=v_\alpha$ and $v_\beta\in Y_y$ for some $\alpha\in\{0,\ldots,r\}$ and $\beta\in\{1,\ldots,r-1\}$.  Without loss of generality, $\alpha < \beta$.
%
%     For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part of $\mathcal{Y}$ that contains $v_i$.  Since $G[\{v_\alpha,\ldots,v_\beta\}]$ is connected, $H[\{x_\alpha,\ldots,x_\beta\}]$ connected.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, $x_j$ is a $T$-ancestor of every node in $x_\alpha,\ldots,x_\beta$ for some $j\in\{\alpha,\ldots,\beta\}$.  In particular $x_j$ is a $T$-ancestor of $x=x_\alpha$ and $y=x_\beta$, so $x_j$ is a $T$-ancestor of $a(v)$. Since $v_j\in Y_{x_j}$, and $v\not\in Y_{a(v)}$ this implies that $v$ participates in $Y_{x_j}$.  Therefore $v$ participates in a $T$-ancestor of $a(v)$, which means that $v$ participates in $a(v)$.
%
%
%   \end{compactenum}
% \end{proof}

% We let $X_v$ denote the set of nodes $x\in V(T)$ such that $v$ participates in $x$.

% \begin{clm}\label{x-v-ancestor}
%   For any $v\in V(G)$ there exists a (unique) node $a(v)\in X_v$ such that
%   $a(v)$ is a $T$-ancestor of every node in $X_v$.
% \end{clm}
%
% \begin{proof}
%   Let $Z := \{v\} \cup \{\{v_1,\ldots,v_{r-1}\}:(v_0,\ldots,v_r)\in\SS, v\in \{v_0,\ldots,v_r\}\}$. Then $G[Z]$ is connected because $Z$ is the union of (vertex sets of) paths in $G$, each of which contains $v$.
%
%   We claim that $v$ participates in a node $x\in V(T)$ if and only if $Z\cap Y_x\neq\emptyset$.  If $v$ participates in $x$ then either $v\in Y_x$, so $Z\cap Y_x\supseteq\{v\}$; or $v\in \{v_0,\ldots,v_r\}$ for some shortcut $(v_0,\ldots,v_r)\in\SS$ that crosses $x$, so $Z\cap Y_x\supseteq \{v_i\}$ for some $i\in\{1,\ldots,r-1\}$.  In the other direction, if $Z\cap Y_x\neq\emptyset$, then either $Z\cap Y_x\supseteq \{v\}$, so $v\in Y_x$; or $Z\cap Y_x\supseteq \{v_i\}$ where $i\in\{1,\ldots,r\}$, $(v_0,\ldots,v_r)\in\SS$ and $v\in\{v_0,\ldots,v_r\}$, so $v\in V(P)$ for a path $P=(v_0,\ldots,v_r)\in\SS$ that crosses $x$.
%
%
%   Let $X_H:=\{x\in V(H): Z\cap Y_x\neq\emptyset\}$.  The connectivity of $G[Z]$ implies that $H[X_H]$ is connected.  This implies that $T':=T[]
%
%
%
%   %Property~\tref{ancestor-edge} implies that, for every $xy\in E(H)$, one of $x$ or $y$ is a $T$-ancestor of the other.
%   Choose $a(v)\in X_H$ to be the member of $X_H$ that does not have a strict $T$-ancestor in $X_H$.  Transitivity of the $T$-ancestor relationship, \tref{ancestor-edge}, and connectivity of $H[X_H]$ implies that such an $a(v)$ exists and is a $T$-ancestor of every node $x\in X_H$, as required.
% \end{proof}

% \note{DW}{Use $\SS$ for the given shortcut system and $\PP$ for the partition?}
% \note{PM}{Yes, I've done this.  Again, I was carefuly, but need to reread.}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\PP:=(S_x : x\in V(T))$ is a sequence of pairwise disjoint sets that cover $V(G)$. Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

% \referee{2}{Page 8.  Can you add a brief explanation why $\PP$ is a
% 	partition of $V(G)$?  It is clear that the sets in $\PP$ are
% 	disjoint, but why do they cover $V(G)$?}
% \note{PM}{I added an explanation in a footnote since this is obvious.  In particular, for any function $f$, $\{f^{-1}(x):x\in\operatorname{range}(f)\}$ is a partition of $\operatorname{domain}(f)$.}

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\PP$ has small layered width with respect to the layering $\mathcal{L}$ of $G$ and that
  \item $J$ has small treewidth.
\end{compactenum}
Once we have established (i) and (ii), the result follows easily since a layering of $G^\SS$ is easily obtained from $\mathcal{L}$ by `compressing' groups of $k$ consecutive layers.
% \referee{1}{7. page 8, item (i): ``$P$ has small layered width with respect to the layering $L$'' $\longrightarrow$ ``$P$ has small layered width with respect to the layering $L$ of $G$''.}
% \referee{1}{Perhaps mention something about how it is easy to find the new layering.}

% First we need to understand the relationship between sets $S_x\in\PP$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

% \referee{1}{8. general comment: Claims 1, 2, and 4 are especially belaboured by
% notation as they are quite straightforward.}

% \note{PM}{Same comment: Yes, these claims are definition checking.  We'll see if the experimental notation change helps.}

% \begin{clm}
% 	\label{s-subset}
%   For every $x\in V(T)$,
%   % $S_x\subseteq V_x$.
%   $S_x\subseteq \bigcup_{x'\in V(T_x)} Y_{x'}$.
% \end{clm}
%
% \begin{proof}
%   Let $v$ be any vertex in $S_x$ (which implies that $a(v)=x$) and let $Y_y$ be the part of $\mathcal{Y}$ that contains $v$.  By definition, $v$ participates in $y$ so by definition $a(v)$ is a $T$-ancestor of $y$, so $y\in V(T_{a(v)})$.  Therefore, $v\in Y_y\subseteq \bigcup_{x'\in V(T_x)} Y_{x'}$, as required.
% \end{proof}

% Next we complete Step~(i) and show that $\PP$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

  Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\SS$ includes $w$ as an internal vertex then $w$ contributes at most one vertex, itself, to $S_x\cap L_i$.
  Otherwise, consider some path $P\in\SS$ that contains $w$ as an internal vertex.  If $w\in L_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap L_i$.  If $w\in L_{i-1}\cup L_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap L_i$. If $w\in L_{i-j}\cup L_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap L_i$.
%
% \referee{1}{9. page 8, line 10: ``We say that a vertex $w \in Y_x$ contributes a vertex
%   $v \in S_x$ if $v$ participates in $x$.'' This needs to be fixed.}
%
% \note{DW}{What is wrong here?}

@@

We now proceed with Step~(ii), showing that $J$ has small treewidth. To accomplish this, we construct a small width tree-decomposition $\mathcal{C}:=(C_x:x\in V(T))$ of $J$ using the same tree $T$ used in $\mathcal{T}$.  The following claim will be useful in showing that the resulting decomposition has small width.

@@

\begin{proof}
  If $xy\in E(J)$ then there exists a shortcut $P:=v_0,\ldots,v_r\in\SS$ with endpoints $v_0\in S_x$ and $v_r\in S_y$.\footnote{Recall that we have made the assumption that $\SS$ contains a length-$1$ $vw$-shortcut for each edge $vw\in E(G)$.}  For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part in $\mathcal{T}$ that contains $v_i$.  Since $P$ is a connected subgraph of $G$, $H[\{x_0,\ldots,x_r\}]$ is a connected subgraph of $H$.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, some node $x_i\in V(T)$ is a $T$-ancestor of all nodes in $x_0,\ldots,x_r$.

@@

  Suppose, without loss of generality, that $v_0$ participates in $x_i$.  Then $a(v_0)=x$ is a $T$-ancestor of $x_i$ which a $T$-ancestor of $x_r$.  Finally $a(v_r)=y$ is a $T$-ancestor of $x_r$.  Therefore, both $x$ and $y$ are contained in the path from the root of $T$ to $x_r$, so one is a $T$-ancestor of the other.
  %
  %
  % is a lazy walk in $H$.
  %
  % Let $P:=v_0,\ldots,v_r$ (where $v_0=v$ and $v_r=w$)
  % Suppose, for the sake of contradiction that neither $x$ nor $y$ is an ancestor of the other.  Then
  %
  %
  %
  %
  %
  %
  % If $P$ has length $1$ (so $vw\in E(G)$) then $
  %
  %
  %
  % By \cref{s-subset}, $v\in G_x$ and $w\in G_y$.
  %
  %
  %
  %
  % Suppose, for the sake of contradiction, that neither $x$ nor $y$ is a $T$-ancestor of the other.  Since $xy\in E(J)$, $G^\SS$ contains an edge $vw$ with $v\in S_x$ and $w\in S_y$.  Since $vw\in E(G^{\SS})$,  $\SS$ contains a $vw$-shortcut $P$.  By \cref{s-subset}, $v\in V_x$ and $w\in V_y$.  By \yref{containment-i}, if neither $x$ nor $y$ is a $T$-ancestor of the other, then $V_x$ and $V_y$ are disjoint.  By \yref{ancestor-edge}, $N_x$ and $V_y$ are also disjoint.  By \yref{separator} $P$ contains an internal vertex $v'\in F_x$.  By \yref{ancestor-edge}, $v'\in Y_a$ for some strict $T$-ancestor $a$ of $x$.  But this implies that $a(v)=a'$ so $v\in S_{a'}$ for some $T$-ancestor $a'$ of $a$, contradicting the assumption that $v\in S_x$.
\end{proof}

@@

% \referee{1}{10. page 10, Claim 5: Indices could mostly be removed by continuing
% to talk about $T$-ancestors instead. Move the definition of $H+$ and its
% directed version up, along with the explanation of what will be proven,
% so that it isn’t necessary to define the $s_i$.}
%
% \note{PM}{I also found the steps in this proof tedious. Unfortunately, when I tried to give a simple explanation for it in Banff, both Piotr and Torsten remained unconvinced until we opened the paper and walked through the steps of the proof together.  I suspect the referee who can see through these proofs if one of the authors of \cite{PS21}.}

>>>>

The purpose of this section is to prove our main technical result, \cref{ShortcutPartition}. This theorem shows how, given a $(k,d)$-shortcut system $\SS$ of a graph $G$, an $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\SS}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.

@@

% \note{PM}{I have significantly simplified this section, which should make the referees happy. I will read it very carefully again tomorrow}

We begin with a standard technique that allows us to work with a normalised tree-decomposition whose tree has the same vertex set as the graph it decomposes:

@@

  Next, remove the node $x_0$ from $T_0$, and consider any node $a\in V(T_0)$ such that there is no vertex $x\in V(H)$ with $f(x)=a$.  In this case, $B_{a}\subseteq B_{a'}$ where $a'$ is the parent of $a$ since any $x\in B_a\setminus B_{a'}$ would have $f(x)=a$.  In this case, contract the edge $aa'$ in $T_0$, eliminating the node $a$.  Repeating this operation will eventually produce a width-$t$ tree-decomposition of $(B_x:x\in V(T_0))$ where $f$ is a bijection between $V(H)$ and $V(T_0)$.  Renaming each node $a\in V(T_0)$ as $f^{-1}(a)$ gives a tree-decomposition $(B_x:x\in V(T))$ with $V(T)=V(H)$.  By the definition of $f$, the tree-decomposition $(B_x:x\in V(T))$ satisfies \tref{subtree-root}.
\end{proof}

@@

@@

\begin{proof}
Let $\mathcal{L}:=\langle L_1,\ldots,L_h\rangle$ be a layering of $G$; let $\mathcal{Y}:=(Y_x: x\in V(H))$ be an $H$-partition of $G$ of layered width at most $\ell$ with respect to $\mathcal{L}$; and let $\mathcal{T}:=(B_x:x\in V(T))$ be a tree-decomposition of $H$ satisfying the conditions of \cref{nice-decomposition}.

For a node $x\in V(T)$, we say that a shortcut $P\in\SS$ \defin{crosses} $x$ if $Y_x$ contains an internal vertex of $P$.  In other words, if $P=v_0,\ldots,v_r$ and $\{v_1,\ldots,v_{r-1}\}\cap Y_x\neq\emptyset$.  We say that a vertex $v\in V(G)$ \defin{participates} in $x$ if $v\in Y_x$ or if $\SS$ contains a shortcut $P$ with $v\in V(P)$ and $P$ crosses $x$.  For each $v\in V(G)$, let $a(v)$ denote the lowest common $T$-ancestor of all the nodes $x\in V(T)$ in which $v$ participates.

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Since $a:V(G)\to V(T)$ is a function with domain $V(G)$ and range $V(T)$, $\PP:=(S_x : x\in V(T))$ is a sequence of pairwise disjoint sets that cover $V(G)$. Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\PP$ has small layered width with respect to the layering $\mathcal{L}$ of $G$ and that
  \item $J$ has small treewidth.
\end{compactenum}
Once we have established (i) and (ii), the result follows easily since a layering of $G^\SS$ is easily obtained from $\mathcal{L}$ by `compressing' groups of $k$ consecutive layers.  We begin with Step~(i):

@@

  Refer to \cref{contribute}.  If $w\in Y_x\cap L_i$ and no path in $\SS$ includes $w$ as an internal vertex then $w$ contributes at most one vertex, itself, to $S_x\cap L_i$.  Otherwise, consider some path $P\in\SS$ that contains $w$ as an internal vertex.  If $w\in L_{i}$, then $P$ contributes at most $k+1$ vertices to $S_x\cap L_i$.  If $w\in L_{i-1}\cup L_{i+1}$, then $P$ contributes at most $k$ vertices to $S_x\cap L_i$. If $w\in L_{i-j}\cup L_{i+j}$ for $j\ge 2$, then $P$ contributes at most $k-j$ vertices to $S_x\cap L_i$.

@@

We now proceed with Step~(ii), showing that $J$ has small treewidth. To accomplish this, we construct a small width tree-decomposition $\mathcal{C}:=(C_x:x\in V(T))$ of $J$ using the same tree $T$ used in the tree decomposition $\mathcal{T}$ of $H$.  The following claim will be useful in showing that the resulting decomposition has small width.

@@

\begin{proof}
  If $xy\in E(J)$ then there exists a shortcut $P:=v_0,\ldots,v_r\in\SS$ with endpoints $v_0\in S_x$ and $v_r\in S_y$.\footnote{Recall that we have made the assumption that $\SS$ contains a length-$1$ $vw$-shortcut for each edge $vw\in E(G)$.}  For each $i\in\{0,\ldots,r\}$, let $Y_{x_i}$ be the part in $\mathcal{Y}$ that contains $v_i$.  Since $P$ is a connected subgraph of $G$, $H[\{x_0,\ldots,x_r\}]$ is a connected subgraph of $H$.  Since $\mathcal{T}$ is a normalized tree decomposition of $H$, some node $x_i$ is a $T$-ancestor of all nodes in $x_0,\ldots,x_r$.

@@

  Suppose, without loss of generality, that $v_0$ participates in $x_i$.  Then $a(v_0)=x$ is a $T$-ancestor of $x_i$ which a $T$-ancestor of $x_r$.  Finally $a(v_r)=y$ is a $T$-ancestor of $x_r$.  Therefore, both $x$ and $y$ are contained in the path from the root of $T$ to $x_r$, so one of $x$ or $y$ is a $T$-ancestor of the other.
\end{proof}

@@

\begin{proof}
  For the tree-decomposition $(C_x:x\in V(T))$ of $J$ we use the same tree $T$ used in the tree-decomposition $(B_x:x\in V(T))$ of $H$. For each node $x$ of $T$, we define $C_x$ as follows: $C_x$ contains $x$ as well as every $T$-ancestor $a$ of $x$ such that $J$ contains an edge $ax'$ where $x$ is a $T$-ancestor of $x'$ (including the possibility that $x=x'$).
  \cref{i-ancestor} ensures that, for every edge $ax'\in E(J)$, both $a$ and $x$ are contained in $C_{x'}$.  The connectivity of $T[a]:=T[\{x\in V(T):a\in C_x\}]$ follows from the fact that, for every node $x'\in V(T[a])$, every node $x$ on the path in $T$ from $a$ to  $x'$ is also a node of $T[a]$.  Therefore $(C_x:x\in V(T))$ is indeed a tree-decomposition of $J$.

  It remains to show that each bag in our tree decomposition has size at most $\binom{k+t}{t}$.  We will do this by appealing to an elegant result of \citet{PS21} on $k$-reachability in directed skeletons, which we now explain.

  Let $H^+$ denote the supergraph of $H$ that contains the edge $xy$ if and only if some bag $B_z\in\mathcal{T}$ contains both $x$ and $y$ (i.e., $H^+$ is the chordal closure of $H$ with respect to $\mathcal{T}$).  Let $\overrightarrow{H}^+$ be the directed graph obtained by directing each edge $xy$ of $H^+$ in the direction $\overrightarrow{xy}$ so that $y$ is a $T$-ancestor of $x$. Observe that directed path in $\overrightarrow{H}^+$ that begins at $x$ leads to a $T$-ancestor of $x$.  \citet[Lemma~13]{PS21} show that, for any $x\in V(\overrightarrow{H}^+)$, the number of strict $T$-ancestors of $x$ that can be reached by a directed path of length at most $k$ that begins at $x$ is at most $\binom{k+t}{t}-1$.
